---
output:
  html_document:
      code_folding: "hide"
---
# TDF Exploration

## Brian Richardson
## 11/28/2022

In this program we do exploratory data analysis on Tour de France data using data obtained from: https://github.com/thomascamminady/LeTourDataSet.

The Tour de France is a multi-stage cycling race that has been held in France annually since 1903, with the exception several canceled tours during World War I and II. Data are available on the stages and riders in each tour from 1903 to 2021. Each tour involves between 6 and 31 stages and has between 10 and 174 riders.

```{r setup, echo=F, message=F, warning=F}

# Prepare workspace
rm(list=ls())

library(ggplot2)
library(tidyr)
library(dplyr)
library(imager)

```

```{r load-data}

### load plots

locations_plot = load.image("figures/locations_plot.png")
riders_per_year_plot = load.image("figures/riders_per_year_plot.png")
riding_time_over_time_plot = load.image("figures/riding_time_over_time_plot.png")
stages_per_year_plot = load.image("figures/stages_per_year_plot.png")
team_wins_plot = load.image("figures/team_wins_plot.png")
rider_wins_plot = load.image("figures/rider_wins_plot.png")
avg_speed_over_time_plot = load.image("figures/avg_speed_over_time_plot.png")
avg_speed_over_time_plot_dope = load.image("figures/avg_speed_over_time_plot_dope.png")
distance_over_time_plot = load.image("figures/distance_over_time_plot.png")

```


The number of stages per year has changed over time. We can investigate this change with the following plot.

```{r stages-plot}

plot(stages_per_year_plot, axes = F)

```

The number of riders per year has changed over time. We can investigate this change with the following plot.

```{r rider-count-plot}

plot(riders_per_year_plot, axes = F)

```

Some starting and ending locations of stages are more common than others. We can look at which locations are most common.

```{r locations-plot}

plot(locations_plot, axes = F)

```

Which teams have won the most tours?

```{r team-wins-plot}

plot(team_wins_plot, axes = F)

```

Which riders have won the most tours?

```{r rider-wins-plot}

plot(rider_wins_plot, axes = F)

```

How has the total distance of each tour changed over time?

```{r distance-plot}

plot(distance_over_time_plot, axes = F)

```

Plot the total riding time by rider for each tour.

```{r riding-time-plot}

plot(riding_time_over_time_plot, axes = F)

```

Plot average speed by rider for each tour.

```{r avg-speed-plot}

plot(avg_speed_over_time_plot, axes = F)
  
```


Take a closer look at the period surrounding Lance Armstrong's illegitimate wins.

```{r avg-speed-plot-dope}

plot(avg_speed_over_time_plot_dope, axes = F)

```



